### 2021-07-06 rad-icfp-2021/unf.tex

We give a translation from our Source to Source UNF in Figure~\ref{fig:source_to_unf}.

TODO:redo

UNF acts as a sort of stack. 
The input represents the current stage of the stack, and we return the new value added to the stack.
All results are kept, so the stack keeps growing. One might be surprised that the transformation for pairs and let is similar.
The order of UNF really follows what the trace of doing a sequential call-by-value evaluation would look like if the return value does not matter.
In other words, when evaluating !let x= e1 in e2! in call by value, we first evaluate !e1! and then !e2!. 
Similarly in !<e1, e2>!, we evaluate !e1! and then !e2!. 
Our ultimate goal is to return a gradient, not the original term, so what really matters is the order of evaluation, and what is evaluated.

TODO: need to explain that return values produce some kind of ANF and introduce new vars.

@@

The transformation from the target UNF to the target language is presented in Figure~\ref{fig:unf_to_target}.

>>>>

We first give a translation from our Source to Source UNF in Figure~\ref{fig:source_to_unf}, which we call $\UNFSymbol$.
As discussed above, $\UNFSymbol$ sends a constant or an operator to a primitive in Source UNF. 
This term in Source UNF carries the context of the original term in Source.
$\UNFSymbol$ sequentializes a term, mimicking a left-to-right call-by-value evaluation.
Because of this, we we're trying to return a pair, this needs to be witnessed, and this is the role of !pair!.
!e!$_{1} \pcomp$ !e!$_{2}$  is used to pass the result !e!$_{1}$ which is at the last position in the tuple through !e!$_{2}$.
It required when the coming operator represents a non-unary primitive such as $+$ or !pair!.
We want to preserve the invariant that $\UNFSymbol$(!e!) represents the term !e! which is also returning its context. 
Because !x! is not free in $\Gamma \vdash$ !let x=e$_1$in e$_2$!, we need to hide !x! after $\UNFSymbol$(!e$_2$!).
This explain the projection !proj! in $\UNFSymbol$ of a !let!. 
More generally, in !e!$_{1} \comp$ !e!$_{2}$ the return value of !e!$_{1}$ acts as a new variable for !e!$_{2}$, 
is not in the context of !e!$_{1}$, but will be returned by !e!$_{2}$.
This means we need to hide all these intermediate values to preserve the invariant,
and this explains the projection !proj! in the $\UNFSymbol$ of other terms.

TODO: we could merge everything of the form op;proj into just op by changing the type of op.

@@

Next, the transformation from Target UNF to Target is presented in Figure~\ref{fig:unf_to_target}. 
We call this transformation $\UNFSymbol^{-1}$, but it is not a strict inverse of $\UNFSymbol$.
Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some kind of ANF. 
TODO: finish
